A free/open-source Kazakh-Tatar machine translation system

### 2013-04-22 paper/paper.tex
headed to Manchu, but now selimcan and spectie should really check §3.4

%FIXME: use a noun in examples, not infinitive/gerund
An example of a difference in verb valencies is with the expression corresponding to ``to like something'' in Kazakh and Tatar.  In Kazakh, the verb ұна is used, e.g. бауырсақ маған ұнайды ``I like bawyrsaq'', where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the object ``bawyrsaq'' in English is the grammatical subject in Kazakh.  Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, e.g. мин бавырсак яратам ``I like bawyrsaq'', where the first person pronoun is the grammatical subject and ``bawyrsaq'' is the grammatical direct object (with no accusative suffix since it is indefinite).
%(does this count as syntactic?)

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\a\label{ex:kerek}
\begingl
\gla Мен үйге қайтуым керек. //
\glb мен үй-GA қайт-у-Iм керек //
\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
\glft \eng{I need to go home.} //
\endgl
\a
\begingl
\gla Миңа өйгә кайтырга кирәк. //
\glb мин-GA өй-GA кайт-IргA кирәк //
\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
\glft \eng{I need to go home.} //
\endgl
\xe

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\a\label{ex:try}
\begingl
\gla Айгүл оны табуға әрекет жасайды. //
\glb Айгүл о-NI тап-у-GA әрекет жаса-E-дI //
\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Aor} //
\glft \eng{Aygül tries to find him.} //
\endgl
\a
\begingl
\gla Айгөл аны табарга тырыша. //
\glb Айгөл а-NI тап-AргA тырыш-E //
\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
\glft \eng{Aygöl is trying to find him.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:GAlI}
\a
\begingl
\gla Мен сенімен сөйлескелі келдім. //
\glb мен сен-і-мен сөйле-с-GAлI кел-DI-м. //
\glb I you-\gmk{Acc}-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\a
\begingl
\gla Мин синең белән сөйлешергә килдем. //
\glb мин син-нIң белән сөйле-ш-IргA кил-DI-м. //
\glb I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe

>>>>

%FIXME: use a noun in examples, not infinitive/gerund
An example of a difference in verb valencies is with the expression corresponding to ``to like to do something'' in Kazakh and Tatar.  In Kazakh, the verb ұна is used, as shown in example (\ref{ex:like-kaz}), where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the gerundive ``writing dictionaries'' is the grammatical subject.  Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, as shown in example (\ref{ex:like-tat}), where the first person pronoun is in nominative case as the grammatical subject and the infinitival verb phrase is the grammatical direct object.
%(does this count as syntactic?)

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\a\label{ex:like-kaz}
\begingl
\gla Маған сөздік түзген ұнайды. //
\glb маған сөздік түз-GAн ұна-E-дI //
\glb \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\a\label{ex:like-tat}
\begingl
\gla Мин сүзлек төзергә яратам. //
\glb мин сүзлек төз-IргA ярат-E-м //
\glb \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:kerek}
\a
\begingl
\gla Мен үйге қайтуым керек. //
\glb мен үй-GA қайт-у-Iм керек //
\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
\glft \eng{I need to go home.} //
\endgl
\a
\begingl
\gla Миңа өйгә кайтырга кирәк. //
\glb мин-GA өй-GA кайт-IргA кирәк //
\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
\glft \eng{I need to go home.} //
\endgl
\xe

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:try}
\a
\begingl
\gla Айгүл оны табуға әрекет жасайды. //
\glb Айгүл о-NI тап-у-GA әрекет жаса-E-дI //
\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Aor} //
\glft \eng{Aygül tries to find him.} //
\endgl
\a
\begingl
\gla Айгөл аны табарга тырыша. //
\glb Айгөл а-NI тап-AргA тырыш-E //
\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
\glft \eng{Aygöl is trying to find him.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:GAlI}
\a
\begingl
\gla Мен сенімен сөйлескелі келдім. //
\glb мен сен-I-мен сөйле-с-GAлI кел-DI-м. //
\glb I you-\gmk{Acc}-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\a
\begingl
\gla Мин синең белән сөйлешергә килдем. //
\glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
\glb I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe
